NumPy---Array Programming in Python

### 2020-01-31 body.tex
minor edits

\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community developed, open source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become widespread as
a language for array programming.

@@

To complement the array syntax, NumPy includes functions that perform
\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
trigonometry.
Vectorization---operating on whole arrays rather than their individual
elements---is essential to array programming.
This means that operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.
This results in concise code and frees users to focus on the details of
their analysis, while NumPy handles looping over array elements near-optimally,
taking into consideration, for example, strides, in order to best utilize the
computer's fast cache memory.

@@

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted usage of
scientific Python.  Examples of its impactful application, such as the
above, now exist in almost every discipline in the natural and social
sciences, and in many fields these tools have become \emph{the
primary} software environment.  NumPy and its ecosystem are commonly
taught in university courses, bootcamps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It runs on machines ranging from embedded devices to the world's
largest supercomputers and yields performance
approaching that of compiled languages.
For algorithms that do not fit the array programming approach,
or require even faster implementations, a variety of tools exist (Cython, Numba, or Pythran, ...) that transparently accelerates bottlenecks.
These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure Python.
% smooth experience/intengration from high level <-> low level?

@@

\begin{figure*}
  \centering
  \includegraphics[width=.9\textwidth]{static/sketches/array_protocol.jpg}
  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API can operate on various types of array objects, including those written outside of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and a dask array, respectively. The call gets succeeds by dispatching to the appropriate library implementation, and results in different output array types.}\label{fig:duck-arrays}
\end{figure*}

@@

Even though NumPy is not part of Python's standard library,
we benefit from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have more succinct and 
easy to read array notation.
Since we are not part of the standard library, we are able to
dictate our own release policies and development patterns.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
Popular array-aware libraries increased the user community and that
in turn led to more libraries adding new functionality
to the scientific Python ecosystem.

>>>>

% KJM: I moved this here to get the placement right 
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community developed, open source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the \emph{de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
\emph{Application Programming Interface} (API) has become widespread as
a language for array programming.

@@

To complement the array syntax, NumPy includes functions that perform
\emph{vectorized} calculations on arrays, including arithmetic, statistics, and
trigonometry (Fig.~\ref{fig:array-concepts}d).
Vectorization---operating on whole arrays rather than their individual
elements---is essential to array programming.
This means that operations that would take many tens of lines to express in
languages such as C can often be implemented as a single, clear Python
expression.
This results in concise code and frees users to focus on the details of
their analysis, while NumPy handles looping over array elements near-optimally,
taking into consideration, for example, strides, in order to best utilize the
computer's fast cache memory.

@@

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted usage of
scientific Python.  Examples of its impactful application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  In many fields these tools have become \emph{the
primary} software environment.  NumPy and its ecosystem are commonly
taught in university courses, bootcamps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It runs on machines ranging from embedded devices to the world's
largest supercomputers and yields performance
approaching that of compiled languages.
For algorithms that do not fit the array programming approach,
or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
that transparently accelerate bottlenecks.
These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure Python.
% smooth experience/intengration from high level <-> low level?

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array_protocol.jpg}

  \caption{\textbf{Interoperability.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{sum} function is called on a NumPy and
     a Dask array, respectively. The call succeeds by dispatching to the
     appropriate library implementation, and results in different output array
     types.
  }\label{fig:duck-arrays}

\end{figure*}

@@

Even though NumPy is not part of Python's standard library,
it benefits from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have more succinct and 
easy to read array notation.
Since it is not part of the standard library, NumPy is able to
dictate its own release policies and development patterns.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
